---
title: "Introduction to Basic Data Types and Objects"
output: html_notebook
---

Today we are going to discuss the foundations of data in R, namely, the basic types of data and objects R supports and manipulates.

## Data types

The basic types of data in R:

1. Numeric:
    i) Integer;
    ii) float or double;
2. Logical;
3. Character (strings);
4. Complex;

```{r types}
2+2
```


These together form the basis of *Atomic* vectors (See the chapter on [vectors](http://r4ds.had.co.nz/vectors.html) in R for Data Science).

## Data objects

There are simple objects:

1. Vectors;
2. Arrays/Matrices

```{r vectors}
# This is an example of a vector
vector_1 <- c(1,2,3,4)

vector_1
```

```{r vector2}
# This is a different, longer vector
vector_2 <- c(1:60)
vector_2
typeof(vector_2)
```

```{r strings}

(string_1 <- c("abc", "def", "ghi"))
typeof(string_1)
length(string_1)

string_2 <- c("abc, def, ghi")
string_2
length(string_2)

long_string <- c(LETTERS)
long_string
```

```{r logicals}
# a vector of true/false

t_f <- c(TRUE, FALSE, T, F)
t_f

c(TRUE, FALSE, 1, 2)
```


```{r factors}
# start with a string
factor_1 <- c("a", "b", "a", "b", "c")
# use coercion to make it a factor
factor_2 <- as.factor(factor_1)
factor_2
```

## Operations

### Basic math operators:
$+, -, *, /$


### MATRIX operators:
%*%  
%+%

```{r simple_operations}
# add two vectors
vector_add <- vector_1 + vector_1

vector_add

vector_multi <- vector_1 * vector_1
vector_multi

# this demonstrates recycling, which is using the same elements over and over
vector_add_2 <- vector_1 + vector_2
vector_add_2
```

```{r arrays}

array_1 <- rbind(vector_1, c(5:8))
array_1

```

```{r}
array_2 <- cbind(vector_1, c(5:8))
array_2
```

```{r dimensions}
length(array_1)
length(array_2)

dim(array_1)
dim(array_2)
```

```{r add-array} 
# arrays have to be the same shape and size
array_1 + array_2
```


## coercion

All vectors can only be **one type**

```{r}

string_3 <- c(1,2,3,4, 4.5, "a")
string_3
vector_4 <- c(1,2, 3.3, 4.8)
vector_4

as.integer(vector_4)

round(vector_4)
```


## Indexing vectors

Vectors and arrays can be indexed to find specific elements

```{r indexing}

length(vector_4)
# find the 3rd element of the vector
vector_4[3]

# find multiple elements using a range
subset_1 <- vector_1[1:3]
subset_1

vector_2[1:10]

vector_1[c(3,2,4)]

vector_2[vector_1]

# multi-dimensional index
array_1[2, 4]
array_1

array_2[4, 2]
array_2

# this shows recycling of the second (column) vector
array_2[c(3,1,4), c(2, 1)]
array_2

```




and more complex objects:

3. Data Frames/Tibbles
    i) See the [tibbles](http://r4ds.had.co.nz/tibbles.html) chapter for more information
4) Lists




